Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-16 submission/manuscript.Rmd
discussion 2-cell and sex

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes. This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

Although expressed at the 2-cell stage an in ESCs, _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].

Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Proper _Otx2_ expression, coupled normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells

Since epiblast stem cells can reactivate germline genes and differentiate into primoridal germ cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].

### males vs females

While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

@@

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells. While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. It is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter. Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.

>>>>

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes. This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression. _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Proper _Otx2_ expression, coupled with normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

While many genes involved in early germline development are common between males and females, some display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

@@

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017],

However, it is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.

 In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter. Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.
